\chapter{Gauge Theories and Lattice Gauge Theories} \label{chap2}

In this chapter we formally introduce the quantum field theory describing
fermions invariant under local group transformations, and introduce the Yang
Mills Lagrangian in \secref{sec:continuum_gauge}.
\Secref{sec:group_intro} gives a brief overview of basic group
theory, introducing the concepts and tools necessary for the thesis. In
\secref{sec:symmetries} we investigate the defining symmetry properties of our
quantum field theory. After this introduction to continuum physics,
\secref{sec:lattice_intro} discretises space-time and puts our theory on a
lattice. In \secrefs{sec:fermion_doubling,sec:scale_setting}, we review the
similarities and differences between the lattice and the continuum, and stress
where additional care has to be taken. Finally, we give a very quick introduction
to numerical evaluation of lattice theories in \secref{sec:numerical_eval}.

For the introduction, standard texts on the subject matter have been consulted.
Among these are introductory volumes on quantum field theory
\citep[e.g.][]{peskin1995introduction,maggiore2004modern}, lattice gauge theory
\citep[e.g.][]{montvay1997quantum,gattringer2009quantum} and group theory
\citep[e.g.][]{georgi1999lie,fulton2013representation}.

\section{The free Dirac Lagrangian} \label{sec:continuum_gauge}

The undoubtedly most important quantity in Quantum Field Theory (\emph{QFT}) and
Statistical Mechanics (\emph{SM}) are the generating functional, named
$\mathcal{Z}$, and the partition function, for the two theories respectively.
In SM, it is defined as
%
\begin{equation} \label{eq:partition}
  \mathcal{Z}_E = \tr e^{-\beta H}
    = \int \prod_i \mathrm{d} \phi_i e^{-\mathcal{S}_E[\phi_i]},
\end{equation}
%
where we have chosen to represent the trace over the Hamiltonian matrix by an
integral over a continuous system, described by a set of field parameters,
$\phi_i$. There is an analogue between the two theories and one can convert from
one to the other by Wick rotating the time coordinate $x^0_E = -i x^0_M$.
Whenever the possibility for confusion exists we use an $M$ to symbolise
Minkowski space, the world of QFT, and an $E$ to symbolise Euclidean space, the
world of SM.

The action, $\mathcal{S}$, is in turn defined by the Lagrangian density
%
\begin{equation} \label{eq:action-def}
  \mathcal{S}[\phi_i] = \int \mathrm{d}^4 x\, \mathcal{L}[\phi_i(x)],
\end{equation}
%
and our theories are uniquely defined by $\mathcal{L}$ and the
properties of the fields $\phi_i$.

Of physical interest are expectation values of Hermitian operators $\widehat{O}$,
which in the path integral formalism can be computed using the partition
function
%
\begin{equation} \label{eq:expectation_value}
  \big\langle \widehat{O}[\phi_i] \big\rangle = \frac{1}{\mathcal{Z}_E} 
    \int \prod_i \mathrm{d} \phi_i \, O[\phi_i]\, e^{-\mathcal{S}_E[\phi_i]},
\end{equation}
%
It is the set of Hermitian operators that represents physical observables.
These can for example be accessed by the introduction of source terms to the
action
%
\begin{equation}
  \mathcal{Z}_E [\,J_i\,] = \int \prod_i \mathrm{d} \phi_i \,
    e^{-\mathcal{S}_E[\phi_i] - \int J_i \phi_i}
\end{equation}
%
with which we can rewrite the expectation value in terms of functional
derivatives
%
\begin{equation}
  \big\langle O[\phi_i] \big\rangle = O\bigg[ \frac{\delta}{\delta J_i} \bigg]
    \log \mathcal{Z}_E \bigg|_{J=0},
\end{equation}
%
hence the name generating functional.

\subsection{The Dirac Lagrangian}

Our starting point will be the Dirac Lagrangian, which reads
%
\begin{equation}
  \mathcal{L} = \bar{\psi}_i \big(i\gamma^{\mu} \partial_{\mu} - m_i\big)
  \psi_i. \label{eq:ldirac}
\end{equation}
%
Here we have introduced the fermionic fields $\psi_i$. To get the right
statistics, these fields have to obey the anti-commutation relations, and
therefore they have to obey the \emph{Grassmann algebra}
%
\begin{align}
  \big\{ \psi_i, \psi_j \big\} &\equiv \psi_i \psi_j + \psi_j \psi_i = 0,\\
  \big\{ \bar{\psi}_i, \bar{\psi}_j \big\} &\equiv \bar{\psi}_i \bar{\psi}_j +
    \bar{\psi}_j \bar{\psi}_i = 0,\\
  \big\{ \bar{\psi}_i, \psi_j \big\} &\equiv \bar{\psi}_i \psi_j + \psi_j
    \bar{\psi}_i = \delta_{ij}.
\end{align}

This Lagrangian describes the dynamics of a set of free, massive fermions, which
by itself is not very exciting. There are however different ways to add
interactions while still maintaining the desired symmetry properties (which we
will discuss later). Depending on the system one is interested in describing,
these include the addition of a self interaction term for the fermions, e.g. a
four point term, $g \big(\bar{\psi}\psi\big)^2$, which results in the NJL model
\citep{Nambu:1961tp}, the introduction of new fields and interactions, and the
localisation of group symmetries which organically induce new fields and
couplings. For the latter approach we need a short overview of group theory,
which we will present next.

\section{A group theory perspective} \label{sec:group_intro}

A group $G$ is a set and a rule for assigning to every (ordered) pair in the set
a new element also in the set
%
\begin{equation}
  \text{For every} \hskip1ex f, g \in G, \hskip1ex\text{then}\hskip1ex
    h = fg \in G \,,
\end{equation}
%
this assignment is called the group product. The group product is associative,
and in every group a unit element ($e$) exists. Every element also has an inverse so
that
%
\begin{equation}
  \forall\hskip.5ex f \in G, \hskip1ex\exists\hskip.5ex f^{-1}
    \hskip1ex:\hskip1ex f^{-1} f = f f^{-1} = e \,.
\end{equation}
%
A representation of a group $G$ on a complex vector space $V$ is a homomorphism
%
\begin{equation}
  D_{R}:\hskip1em G \to \text{GL}(V)
\end{equation}
%
of the group $G$ on the group of automorphisms of $V$. Therefore representations
have to maintain the group structure, meaning
%
\begin{align}
  D_R(e) &= \mathbb{1}_R,\\
  D_R(f) D_R(g) &= D_R(fg). \label{eq:group_prod}
\end{align}
%
In physics we are normally content with a coordinate basis, in which the
representations map to $n$ by $n$ invertible matrices, i.e. $V := \mathbb{C}$,
$\text{GL}(V) := M_{\mathbb{C}}(n,n)$.

\subsection{Lie groups}

If the elements of a group $G$ depends smoothly on some set of parameters,
$g(\alpha_i) \in G$, we call that group a Lie group. Lie groups make up the
theory of continuous transformations, and are therefore of special interest to
physics. The continuous nature of the Lie groups creates a sense of closeness
where two elements that are close to each other can be represented through
differentials, e.g.
%
\begin{equation}
  D_R(g(\delta\alpha_i)) = \mathbb{1}_R + i\, \delta \alpha_i T_{Ri} +
    \mathcal{O}(\delta \alpha^2).
\end{equation}
%
One commonly refers to the $T_{Ri}$ as the generators of the group $G$. Applying
the smoothness condition to the group product, \meqref{eq:group_prod},
we find that the commutator of the generators follow a special relation
%
\begin{equation} \label{eq:group_algebra}
  \big[ T_{Ri}, T_{Rj} \big] = i\, f_{ijk} T_{Rk}.
\end{equation}
%
The structure constants, $f_{ijk}$, are in fact independent of the
representation $R$. They are tremendously important as they summarise the entire
group multiplication law. They are also referred to as the \emph{algebra} of the
group, $\mathfrak{g}$. The generators and the group algebra will come into play
when we later introduce continuous symmetries of our quantum theory of fields.
Using the Jacobi identity, it is possible to see that the structure constants
themselves comply with the multiplication law. We can therefore use the structure
constants as a representation of the algebra $\mathfrak{g}$ called the
\emph{adjoint representation}
%
\begin{equation}
  \big[ T_{\text{adj}\hskip.2ex i} \big]_{jk} \equiv -i\, f_{ijk},
\end{equation}
%
which fulfils \meqref{eq:group_algebra}.

\subsection{Character and character expansion}

The character of a representation is defined as
%
\begin{equation}
  \chi_R(g) = \tr D_R(g).
\end{equation}
%
The characters appear frequently in field theory due to the cyclic properties of
the trace and its ability to facilitate the creation of invariant objects. On
top of this, the characters span a basis of functions that share this cyclic
property, i.e. any function $\rho(f g f^{-1}) = \rho(g)$. Hence
%
\begin{equation}
  \rho(g) = \sum_r \rho_r \chi_r(g),
\end{equation}
%
with
\begin{equation}
  \rho_r = \int_{\mathrlap{g \in G}\hskip1.5ex} \mathrm{d}g \, \chi_r^*(g) \rho(g).
\end{equation}
%
It is common to factor out both the trivial representation ($D_0(g) = 1$), as
well as the dimension of the representation, and write
%
\begin{equation}
  \rho(g) = \rho_0 \Big(1 + \sum_{r \neq 0} d_r \bar{\rho}_r \chi_r(g)\Big),
\end{equation}
%
which we will refer to as the character expansion. We will make use of this in
\secref{sec:char_exp}.

\subsection{Group integrals}

Lastly we need to define integrals over continuous groups. We introduce
the Haar integration measure, which has the following properties
%
\begin{equation}
  \int \mathrm{d}g \, \rho(g) = \int \mathrm{d}g \, \rho(fg) 
    = \int \mathrm{d}g \, \rho(gf) \hskip2ex \forall \hskip1ex f \in G
\end{equation}
%
and the normalisation
%
\begin{equation}
  \int \mathrm{d}g = 1.
\end{equation}
%
Throughout the derivation of the effective theory, we will encounter polynomial
integrals on the form
%
\begin{equation}
  I = \int \mathrm{d} g \, g^n (g^{-1})^m
\end{equation}
%
or more commonly, with a representation
%
\begin{equation} \label{eq:group-integrals-representation}
  I_{i_1,j_1,...,i_n,j_n}^{k_1,l_1,...,k_m,l_m} = \int \mathrm{d} g \,
    D_R(g)_{i_1j_1} \cdots D_R(g)_{i_nj_n} D_R(g^{-1})_{k_1l_1} \cdots D_R(g^{-1})_{k_ml_m}.
\end{equation}
%
There are multiple ways of handling these integrals, depending on the Lie group
one is interested in. For SU($N$) one can for instance decompose the integral
into generalised Euler angles \citep[as described in][]{Tilma:2004kp}, construct
generating functionals for the integrals
$I_{i_1,j_1,...,i_n,j_n}^{k_1,l_1,...,k_m,l_m}$ \citep[details
in][]{Creutz:1978ub}, or one can use tensor product decompositions to write the
integrals in terms of Young projectors \citep{Myers:2014dia}. In the present
work, a combination of the first two approaches is used, and extensive details
are provided in
\apxrefs{sec:haar_measure,sec:character_integrals,sec:sun_integrals}.

\section{Symmetries of the Lagrangian} \label{sec:symmetries}

With the necessary formalism in place, we can now discuss the symmetries of the
Lagrangian we introduced in \meqref{eq:ldirac}. We will first look at the
symmetries that come by construction before delving into possible ways of
extending the theory to be both globally and locally symmetric under additional
transformations.

\subsection{Symmetries by construction}

Every quantum field theory is constructed to be symmetric under the full
Poincar\'e group, namely the group of space-time translations plus the Lorentz
transformations. 

To see the other symmetries of the Dirac Lagrangian we have to decompose the
fields into its spinorial components. Every element of $\psi_i$ in
\meqref{eq:ldirac} is a four-component spinor, and the $\gamma$'s, $4 \times 4$
matrices. In the standard representation they take the form
%
\begin{equation}
  \psi = \frac{1}{\sqrt{2}} \begin{pmatrix} \psi_R + \psi_L \\ \psi_R - \psi_L \end{pmatrix}, \hskip1cm
    \gamma^0 = \begin{pmatrix} 1 & 0 \\ 0 & -1 \end{pmatrix}, \hskip1ex
    \gamma^i = \begin{pmatrix} 0 & \sigma^i \\ -\sigma^i & 0 \end{pmatrix}.
\end{equation}
%
The fields $\psi_{R,L}$ are two component fields that transform under two
distinct representations of the Lorentz group, $(\frac{1}{2}, 0)$ and $(0,
\frac{1}{2})$ respectively. With this choice of basis the Dirac Lagrangian reads
%
\begin{equation}
  \mathcal{L} = i \bar{\psi}_R \gamma^{\mu}\partial_{\mu} \psi_R
    + i \bar{\psi}_L \gamma^{\mu}\partial_{\mu} \psi_L
    - m (\bar{\psi}_R\psi_L + \bar{\psi}_L\psi_R).
\end{equation}
%
It is apparent that this Lagrangian is symmetric under the U($1$) $\otimes$ U($1$)
transformation
%
\begin{equation} \label{eq:chiral_transformation}
  \psi_{R,L} \to e^{i \alpha_{R,L}} \psi_{R,L}
\end{equation}
%
if and only if the mass term is set to zero. Invariance of the independent
transformations of the left- and right-handed fields is called \emph{chiral}
symmetry, and signals the existence of helicity eigenstates. The presence of the
mass term explicitly breaks the symmetry into a single U($1$) symmetry, where
$\alpha_R = \alpha_L$.

The global symmetry group of $\psi$ can be trivially extended to admit any Lie
group by making the field live in the vector space the representation of choice
acts on, and carry out transformations in the following way
%
\begin{equation} \label{eq:global_trafo}
  \psi \to D_R(g) \psi, \hskip1em \bar{\psi} \to \bar{\psi} D_R^{\dagger}(g),
  \hskip1em g \in G.
\end{equation}
%
This symmetry has an associated conserved Noether current for every generator of
the group $G$
%
\begin{equation}
  j^{\mu}_i = \bar{\psi} \gamma^{\mu} T_i \psi, \hskip .5cm\text{where}\hskip .5cm
    \partial_{\mu}j^{\mu}_i = 0,
\end{equation}
%
which gives the associated charge
%
\begin{equation} \label{eq:noether-charge}
  Q_i = \int \mathrm{d}^3 x\, \bar{\psi} \gamma^0 T_i \psi.
\end{equation}

\subsection{Gauge symmetries}

We just saw that the Lagrangian can be trivially extended to be invariant under
global symmetry transformations. Next, we want to promote them to local
symmetries. This is motivated post facto by QED, which exhibits a local U($1$)
symmetry.

The idea was first explored by Weyl, who tried introducing local scale
transformations to the metric
%
\begin{equation}
  g_{\mu\nu}(x) \to e^{\sigma(x)}g_{\mu\nu}(x)
\end{equation}
%
which prompted the need for an additional vector field to gauge how much one
needs to adjust scales to compare two separate points. Weyl then showed that
these vector fields need to follow Maxwell's equations, and postulated that
the vector fields were in fact nothing else than Maxwell's electromagnetic
potentials \citep{Weyl:1918ib}. It was later discovered that classically scale
invariant theories loose this symmetry due to a quantum effect sometimes
referred to as the \emph{Weyl anomaly}. In fact, it is this very effect that is
responsible for the bulk contribution to the hadron masses in QCD. So, although
Weyl's ideas does not describe how the world actually works, it served as a
cornerstone on which the theories connecting the geometry of space to local
symmetries was built. Instead of adjusting local length scales, we will apply
group transformations to vector spaces that live on top of space-time.

We define a local group element by a Lie group whose parameters are space-time
dependent $g(\alpha_i(x))$. We will use the short hand notation $U(x) =
D_R\big(g(\alpha_i(x))\big)$ for representations of such group elements. We
suppress the representation $R$-subscript wherever it is not needed. The
transformations of the fields at a single coordinate follow trivially from
\meqref{eq:global_trafo}
%
\begin{equation} \label{eq:local_trafo}
  \psi(x) \to U(x) \psi(x), \hskip1em \bar{\psi}(x) \to \bar{\psi}(x)
  U^{\dagger}(x).
\end{equation}
%
There is however no way to make derivative terms invariant under these
symmetries. This becomes obvious when writing the derivative in terms of limits
%
\begin{equation}
  \eta^{\mu} \partial_{\mu} \psi(x) = \lim_{\delta \to 0} \frac{1}{\delta} \big(
    \psi(x + \delta\eta) - \psi(x) \big)
\end{equation}
%
where $\eta^{\mu}$ is some unit vector. We have to take the difference between
fields at two different points in space-time, which transform independently
under \meqref{eq:local_trafo}. Naively calculating this difference is
nonsensical as one is allowed to choose the basis of the representation of the
different points independently, meaning that $\psi(x)$ and $\psi(y)$ can be
members of different vector spaces. We therefore introduce the parallel
transporters, $\Lambda(y,x)$, which transforms between the vector spaces of the
fields at the points $x$ and $y$ in a smooth way \citep{Wu:1975es}. We define
the \emph{covariant derivative} as
%
\begin{equation}
  \eta^{\mu} D_{\mu} \psi(x) = \lim_{\delta \to 0} \frac{1}{\delta} \big(
    \psi(x + \delta\eta) - \Lambda(x+\delta\eta,x)\psi(x) \big).
\end{equation}
%
where the transporter $\Lambda$ make sure we subtract quantities which are
defined on the same vector space. These transformations are nothing but
coordinate transformations of the vector spaces, and must therefore itself be a
member of the representations on G. From this it follows that it transforms as
%
\begin{equation}
  \Lambda(y,x) \to U(y) \Lambda(y,x) U^{\dagger}(x).
\end{equation}
%
This in turn gives rise to the subsequent transformation rule for the covariant
derivative
%
\begin{equation}
  D_{\mu}\psi(x) \to U(x) D_{\mu} \psi(x),
\end{equation}
%
ensuring that terms of the form $\bar{\psi} D_{\mu} \psi$ remain invariant under
local symmetry transformations. As $\Lambda$ is a representation of $G$, we can
express it in terms of the generators of the group. On infinitesimal
form, this is
%
\begin{equation}
  \Lambda(x + \delta\eta,x) = 1 + i g \delta \eta^{\mu} A^i_{\mu}(x) T_i +
    \mathcal{O}(\delta^2)
\end{equation}
%
where the $A^i_{\mu}$ have to transform as
%
\begin{equation}
  A^i_{\mu} \to A^i_{\mu} + \partial_{\mu} \alpha^i - g f^{ijk} \alpha^j A^k_{\mu}
    + \mathcal{O}(\alpha^2),
\end{equation}
%
which is independent of the representation. The local symmetric Dirac
Lagrangian thus reads
%
\begin{equation}
  \mathcal{L} = \bar{\psi}\big(i \gamma^{\mu} D_{\mu} - m\big) \psi.
\end{equation}
%
Unfortunately, without a kinetic term for the newly introduced fields
$A^i_{\mu}$, the theory would permit violently oscillating fields with no cost
in energy \citep{tHooft:2002zz}. This would make our theory non-renormalisable,
which we obviously want to avoid.  We therefore need to find a kinetic term for
the fields that admits all of our restrictions. When constructing such a term,
it is natural to once more turn to the parallel transporters, it being the main
ingredient in adjusting the kinetic term of the fermions. Using the gauge
transformation of $\Lambda$, it is easy to see that the geometric plaquette
could create such a term
%
\begin{align} \label{eq:transport_plaquette}
  U_{\mu\nu}(x) =& \Lambda(x, x + \hat{\nu} \delta)
    \Lambda(x + \hat{\nu} \delta, x + \hat{\mu} \delta + \hat{\nu} \delta) \nonumber\\
    &\hskip1em\times\Lambda(x + \hat{\mu} \delta + \hat{\nu}\delta, x + \hat{\mu} \delta)
    \Lambda(x + \hat{\mu}\delta, x)
\end{align}
%
which transforms as
%
\begin{equation}
  U_{\mu\nu}(x) \to U(x) U_{\mu\nu}(x) U^{\dagger}(x) \,,
\end{equation}
%
and hence $\tr U_{\mu\nu}(x)$ is invariant under local group
transformations. Expanding $U_{\mu\nu}$ in the fields $A^i_{\mu}$, we get
%
\begin{equation}
  U_{\mu\nu} = 1 + i g \delta^2 T_i \big( \partial_{\mu} A^i_{\nu} - \partial_{\nu}
  A^i_{\mu} + g f^{ijk} A^j_{\mu} A^k_{\nu} \big) + \mathcal{O}(\delta^3)
\end{equation}
%
from which we extract the term
%
\begin{equation} \label{eq:field_strength}
  F_{\mu\nu}^i = \partial_{\mu} A^i_{\nu} - \partial_{\nu}
    A^i_{\mu} + g f^{ijk} A^j_{\mu} A^k_{\nu} \,,
\end{equation}
%
which is both invariant under the gauge transformations as well as being the
kinetic term we were looking for. The full gauge extended Lagrangian thus
reads
%
\begin{equation} \label{eq:gauge_lagrangian}
  \mathcal{L} = \bar{\psi}\big(i \gamma^{\mu} D_{\mu} - m\big)\psi 
    - \frac{1}{4} F^i_{\mu\nu} F^{i\hskip.2ex\mu\nu} \,.
\end{equation}
%
There are higher order combinations of these ingredients which are invariant
under the symmetries discussed so far, such as the dual field strength
%
\begin{equation}
  \mathcal{L}_{\mathrm{D}} = g_{\mathrm{D}} \varepsilon^{\alpha\beta\mu\nu}
    F^i_{\alpha\beta} F^i_{\mu\nu}\,,
\end{equation}
%
and magnetic dipole interactions
%
\begin{equation}
  \mathcal{L}_{\mathrm{MD}} = g_{\mathrm{MD}} \bar{\psi} \sigma^{\mu\nu} F^i_{\mu\nu} \psi\,.
\end{equation}
%
We ignore these in the present study as they contribute to the breaking of other
symmetries and/or systematics, some of which we are interested in conserving at
the scope of this thesis, such as CP symmetry and renormalisability,

\section{Lattice discretisation} \label{sec:lattice_intro}

With the Lagrangian at hand we want to actually evaluate the partition function
integral from \meqref{eq:partition}. The integral is however infinite
dimensional, and has no known closed analytical form unless the theory is
trivial (non-interacting). If the theory is close to being trivial, one can
apply perturbation theory and systematically solve the integral order by order.
This is however not the case for the theories we plan to study, and we must
resort to other means.

In the lattice approach one discretises space-time into a grid, or lattice,
restricting the fields and variables to exist only at these points. We thus
introduce a shortest distance, namely the spacing between these grid points,
commonly named $a$. The coordinates are thus restricted to
%
\begin{equation}
  \big(x_0, x_1, x_2, x_3\big) \in \mathbb{R}^4 
  \hskip1ex\to\hskip1ex \big(n_0, n_1, n_2, n_3\big) \in
    \raisebox{.1ex}{\scalebox{0.9}{$\big\{0, 1, ..., N-1\big\}^4$}}.
\end{equation}
%
In addition to having a shortest distance, which works as a natural regulariser
for this approach, we also have a finite volume $\Omega = a^4 N^4$ \footnote{We
  later distinguish between spatial and temporal lattice directions, and denote
  the total number of lattice points in one such direction by $N_s$ and $N_t$.}.
Therefore, to recover physical results, we need to take the continuum limit, $a
\to 0$, and the infinite volume limit, $N \to 0$, in such a way that $\Omega$ is
large enough in physical units to encompass the scales we are interested in
studying.

Having introduced the lattice, we see that the partition function of the
discretised theory is simplified compared to the continuum. The simplification
is not obvious as it quietly changes the number of integrals from being infinite
to being an albeit still large, but number.

In the following sections we will first construct a Lagrangian to put on the
lattice and analyse its new symmetry properties. We will then briefly discuss
how to determine the lattice scale and compare lattice results with continuum
physics. Finally a brief overview of the real world computations and
applications of lattice gauge theories will be given.

\subsection{The lattice gauge theory Lagrangian}

Constructing a \emph{good} discretised Lagrangian is both a subject and an art
of itself. Naively one would think that any Lagrangian that reduces to the
continuum Lagrangian \eqref{eq:gauge_lagrangian} when we send $a \to 0$ would
suffice. Or broader, Lagrangians that at most introduce total derivatives with 
vanishing borders when $a \to 0$. Though this is to some extent true, we will
see that as always, the devil are in the details, and if one does not take great
care, the continuum theory will not be as it first appears.

In the previous section we introduced the parallel transporters to define how to
associate two different points in space that can belong to different vector
spaces. This procedure complies nicely with the lattice approach, as the only modification we
have to make to the definition of the covariant derivative is to
replace $\delta$ by $a$, and abolish the limit. We will exclusively use the
transporter between nearest neighbour lattice sites, and therefore make the
abbreviation $U(x + a \eta_\mu, x) = U_{\mu}(x) =
U_{-\mu}^{\dagger}(x+a\eta_{\mu})$. Hence, the fermion contribution to the
lattice action with a symmetrised derivative reads
%
\begin{equation} \label{eq:naive_fermions}
  \mathcal{S}_f = a^4 \sum_x \Big( 
    m \bar{\psi}(x)\psi(x) - 
    \frac{1}{2a}\sum_{\mu=\pm 0}^{\pm 3}
    \bar{\psi}(x+\hat{\mu})\gamma^{\mu}U_{\mu}(x) \psi(x) \bigg),
\end{equation}
%
where the sum is over the discrete points on the lattice, and we have used the
shorthand definition $\gamma^{-\mu} \equiv -\gamma^{\mu}$. We have also switched
to Euclidean space, which is a conscious choice that will benefit us later in
the stochastic evaluation of the path integral. It is common to factor the
fields out of the definition and write the action in terms of a vector-matrix
product
%
\begin{equation} \label{eq:quark_mat_def}
  \mathcal{S}_f = a^4 \bar{\psi}_y Q_{yx} \psi_x, \hskip1em 
  Q_{yx}[U_{\mu}] = m \delta_{y,x} - \frac{1}{2a} \sum_{\mu=\pm 0}^{\pm 3} \gamma^{\mu}
    U_{\mu}(x) \delta_{y-\mu,x} \,,
\end{equation}
%
where the spin and gauge indices are hidden.  This is to simplify the expression
for the integral over the fermion fields, which can be evaluated exactly
%
\begin{equation} \label{eq:fermion-integral}
  \int \prod_{i,j} \mathrm{d} \bar{\psi}_i \psi_j e^{\bar{\psi}_i Q_{ij} \psi_j}
   = \det Q,
\end{equation}
%
with $Q$ commonly referred to as the \emph{Dirac operator} or the \emph{fermion
  matrix}, and $\det Q$ the \emph{fermion determinant}. The matrix can either be
represented as a higher dimensional tensor, or using a super index describing all
degrees of freedom.

Having introduced the \emph{plaquette} transporter in
\meqref{eq:transport_plaquette}, constructing a lattice gauge action is
straightforward
%
\begin{equation}
  \mathcal{S}_g = \beta_R a^4 \sum_x \sum_{\mu < \nu} \Big( 1 -
    \textstyle\frac{1}{d_R} \hskip.2ex \text{Re} \tr U_{R \hskip.2ex\mu\nu}(x) \Big),
\end{equation}
%
where $d_R$ is the dimension of the representation, $d_R = \tr D_R(e)$. $\beta_R$ is the
coupling constant, and by comparison with the continuum field strength tensor
\eqref{eq:field_strength}, we find
%
\begin{equation}
  \beta_R = \frac{2 d_R}{g^2} \,.
\end{equation}

\section{Fermion doubling and chiral symmetry} \label{sec:fermion_doubling}

Although the Lagrangian \meqref{eq:naive_fermions} has the desired
transformation properties, and reproduces the Dirac Lagrangian in the continuum, it
suffers from a degeneracy in its energy eigenvalues. The easiest way to see this
is by choosing the trivial representation of the gauge fields and writing down
the fermion Hamiltonian
%
\begin{equation}
  \mathcal{H}_f = \sum_{\vec{x}} \bigg(
    m \bar{\psi}(\vec{x}) \psi(\vec{x}) + \frac{i}{2a} \sum_{k=1}^3 \Big(
      \bar{\psi}(\vec{x}+\hat{k})\gamma^k\psi(\vec{x}) 
      -\bar{\psi}(\vec{x}-\hat{k})\gamma^k\psi(\vec{x})\Big)\bigg).
\end{equation}
%
From experience we know that the free Hamiltonian is diagonal in momentum space,
so we introduce a Fourier transform
%
\begin{equation}
  \tilde{\psi}(\vec{p}) = a^3 \sum_{\vec{x}} e^{-i \vec{p}\cdot \vec{x}}
    \psi(\vec{x})
\end{equation}
%
where the possible values of $\vec{p}$ are restricted to the Brillouin zones
%
\begin{equation}
  p_i = \frac{2\pi}{a N} \nu_i, \hskip1ex \nu_i \in
    \big\{\textstyle\minus{}\big(\frac{L}{2}\minus{}1\big), ...,
    \frac{L}{2} \big\} \hskip1ex\Rightarrow\hskip1ex p_i \in \big(\minus\pi, \pi\,\big].
\end{equation}
%
After a Fourier transformation the Hamiltonian is
%
\begin{equation}
  \mathcal{H}_f = \frac{1}{(aN)^3}\sum_{\vec{p}} \tilde{\bar{\psi}}(\vec{p})
    \Big( m + \frac{1}{a} \gamma^k \sin (a p_k) \Big) \tilde{\psi}(\vec{p}).
\end{equation}
%
It can easily be seen that it has eigenvalues
%
\begin{equation}
  H_{\vec{p}} = \pm \sqrt{m^2 + \frac{1}{a^2} \sum_{k=1}^3 \sin^2 (a p_k)}
\end{equation}
%
which for every momentum $\vec{p}$ is $8$-fold degenerate, owing to the
periodicity of the sine function. This implies that we have $7$ additional,
unwanted fermions in the theory whenever we put one in. These \emph{doublers}
come from the corners of the Brillouin zone, and therefore disappear as we go
to the continuum limit. This is however of little comfort, as they would give
incalculable effects to loop diagrams at finite volumes which cannot easily be
factored out. In four dimensional space-time an additional set of $8$ energy
eigenstates appear from doublings in the additional temporal Brillion corners.

\subsection{Wilson fermions}

In an attempt to resolve this issue, Wilson introduced an additional term to the
Lagrangian in the form of a second order derivative. This new term,
appropriately named the \emph{Wilson term}, is one order higher in $a$, and
therefore disappears in the continuum limit. After reordering the terms, adding
the Wilson term to the Lagrangian gives us
%
\begin{equation} \label{eq:wilson_action}
  \mathcal{S}_{f,w} = a^4 \sum_{x} \bigg(
  \big(m + {\textstyle\frac{4 r}{a}}\big) \bar{\psi}(x) \psi(x)
  - \frac{1}{2a} \sum_{\mu = \pm 0}^{\pm 3} \bar{\psi}(x+\hat{\mu}) (r +
  \gamma^{\mu}) U_{\mu}(x) \psi(x) \bigg).
\end{equation}
%
We have introduced the \emph{Wilson parameter} $r$, which has to be in the
interval $(0, 1]$, and is commonly chosen to be $1$. From this one can calculate
the Hamiltonian's eigenvalues, and show that they are modified to
%
\begin{equation}
  H_{\vec{p}} = \pm \sqrt{\bigg(m + \frac{r}{a}\sum_{k=1}^3 (1 - \cos (a p_k) ) \bigg)^2
    + \frac{1}{a^2} \sum_{k=1}^3 \sin^2 (a p_k)}.
\end{equation}
%
The addition of the Wilson term has given an additional mass of $n r/a$ to the
unwanted doublers, meaning that they properly decouple from the theory in the
continuum where their masses diverge. 

However, including the Wilson term does not come without a cost. By introducing
the Wilson term, we explicitly break chiral symmetry. As a result, regardless of
whether we choose $m=0$ for the bare mass parameter, the transformation
\eqref{eq:chiral_transformation} is no longer a symmetry of the system. 

There is in fact a theorem by \citeauthor{Nielsen:1980rz}
[\citeyear{Nielsen:1980rz,Nielsen:1981xu}] which states that one cannot have a
lattice theory both chirally symmetric and free of doublers at the same time
with a Dirac Lagrangian that is also local and possessing the correct continuum
limit.

Yet all hope is not lost as Ginsparg and Wilson shortly after stated a modified
equation the fermion propagator has to fulfil to give a chiral theory in the
continuum \citep{Ginsparg:1981bj}. Although Wilson fermions do not meet these
requirements, solutions to the equations have since been found. Two notable
discretisation schemes are \emph{overlap fermions} \citep{Neuberger:1997fp,Neuberger:1998wv},
\emph{domain wall fermions} \citep{Kaplan:1992bt} (which heed the equation in a
special limit of its internals).

\section{Scale setting and the continuum limit} \label{sec:scale_setting}

We previously mentioned that the lattice spacing $a$ works as a non-perturbative
regulariser of the theory, and thus sets the internal scale. Any parameter in
the lattice theory appears as dimensionless combinations, such as $\vec{x}/a$,
$a \vec{p}$, $a m$, $a^{3/2} \psi(x)$. The scale is therefore to some extent
hidden, and the only way to access it is by computing observables and comparing
them to experimental values
%
\begin{equation}
  \lim_{a \to 0} O(m, g; a) = O_{\text{phys}}.
\end{equation}
%
Simply sending $a$ to $0$ is considered to be the \emph{naive} continuum limit,
as it fails to take into account the parameters' dependence on the scale. The
parameters that enter our theory are called \emph{bare}, as they vary together
with the scale
%
\begin{equation}
  m_{\text{phys}} = \lim_{a \to 0} m_{\text{bare}}(a), \hskip2em
  g_{\text{phys}} = \lim_{a \to 0} g_{\text{bare}}(a).
\end{equation}
%
The scale dependence of the theory's parameters is summarised by the
Callan-Symanzik equations, which can be computed for any physical observable
$O_{\text{phys}}$
%
\begin{equation}
  a \frac{\mathrm{d}}{\mathrm{d} a} O_{\text{phys}}
    = \Big( a \frac{\partial}{\partial a} 
      - \beta \frac{\partial}{\partial g}
      - \gamma \frac{\partial}{\partial m}
    \Big) O_{\text{phys}} = 0.
\end{equation}
%
The $\beta$-function describes how much the coupling changes as a function of
the lattice spacing, while the anomalous dimension $\gamma$ gives the field
strength scaling. To lowest order in perturbation theory, the $\beta$-function
is
%
\begin{align}
  \beta(g) = - a \frac{\partial g}{\partial a} &= - \frac{g^3}{(4\pi)^2}
    \Big( \frac{11}{3} C_{\text{adj}\hskip.2ex 2} - \frac{4}{3} n_f C_R \Big) + \mathcal{O}(g^5), \nonumber\\
  &\equiv - \beta_0 g^3 + \mathcal{O}(g^5) \label{eq:beta_eq}
\end{align}
%
Here we encounter the group's Casmir operators, $C_R$ and $C_{\text{adj}\hskip.2ex 2}$. They are
defined as
%
\begin{equation}
  C_{R \hskip.2ex 2} = || T_{Ri} f^{iab} f^{jba} T_{Rj}||, \hskip2ex
    C_R = \frac{d_R}{d_{\text{adj}}} C_{R \hskip.2ex 2},
\end{equation}
%
and take the following values for the fundamental representation of SU($N$)
%
\begin{equation}
  C_F = \frac{1}{2}, \hskip1em C_{\text{adj}\hskip.2ex 2} = N \,.
\end{equation}
%
One can solve the differential equation in \meqref{eq:beta_eq} and get
%
\begin{equation} \label{eq:running-coupling}
  g = \frac{1}{\sqrt{1 - 2\beta_0 \log (a/a_0)}}
    \hskip1em\Leftrightarrow\hskip1em
    a = \frac{1}{\Lambda} \exp\Big\{\frac{-1}{2\beta_0 g^2}\Big\} .
\end{equation}
%
We see that although $g$ was initially dimensionless, we have generated an
energy scale $\Lambda$ through the renormalisation group, a phenomenon known as
\emph{dimensional transmutation}. We also see that the lattice spacing $a$ goes
to zero as $g$ goes to zero (and the lattice coupling $\beta$ goes to infinity).
This is the \emph{true} continuum limit, where one, as opposed to the naive
limit, takes the change of the parameters into account.

\section{Numerical evaluation of lattice gauge theories}
\label{sec:numerical_eval}

Although the lattice discretisation has decreased the number of integrals in the
definition of $\mathcal{Z}$ from infinity to something finite, the number of free
variables is still too large for it to be evaluated, even numerically. The
fermion fields have $2 N_f d_R$ independent degrees of freedom at every
position. The gauge field is a four vector at every space-time point with
$d_{\text{adj}}$ degrees of freedom. Even though the fermion fields can be
integrated out exactly \eqref{eq:fermion-integral}, this still leaves a highly
dimensional integral if one wants to look at anything but miniscule volumes.

The only feasible way to handle these types of integrals is by using approximate
integration schemes, many of which are stochastic in nature. One such algorithm
is \emph{Monte Carlo integration}, which instead of summing over all
values the degrees of freedom can take, as one does in standard numerical
integration, only samples a small number of these configurations. The value of
the integral is then reconstructed using this (small) number of configurations.

For this to work well one has to sample the configuration space in an
optimal way as the integrand is usually sharply peaked around a small set of
very specific configurations. A first step is to do importance sampling, where
one uses the integrand as a probability distribution for the configuration
space, in the case of the partition function
%
\begin{equation}
  P\big([\psi,\bar{\psi}, U_{\mu}]\big) \propto e^{-\mathcal{S}[\psi,\bar{\psi}, U_{\mu}]}.
\end{equation}
%
This is only a viable interpretation as long as the right hand side is real and
positive. We evaluate the fermion integrals according to
\meqref{eq:fermion-integral}, giving us
%
\begin{equation}
  P\big([U_{\mu}]\big) \propto \det Q[U_{\mu}] \, e^{-\mathcal{S}_g[U_{\mu}]}.
\end{equation}
%
The exponent of the gauge action is clearly both real and positive. The fermion
determinant can be shown for Wilson fermions to be real by using the fact that
it is $\gamma^5$-hermitian
%
\begin{equation}
  \gamma^5 Q \gamma^5 = Q^{\dagger}.
\end{equation}
%
It is however not guaranteed to be positive. We can fix this by studying systems
with an even number of pairwise degenerate quarks. The QCD action does not mix
the different quark species, and we therefore get a separate fermion matrix for
every quark flavour
%
\begin{equation}
  \det Q = \prod_f \det Q_f
\end{equation}

Since direct sampling from the distribution is not possible in practice,
one usually constructs a chain of samples, each being related to the previous
through a Markov chain
%
\begin{equation}
  \Big\{
    \big[\psi, \bar{\psi}, U_{\mu}\big]_1, 
    \big[\psi, \bar{\psi}, U_{\mu}\big]_2, ..., 
    \big[\psi, \bar{\psi}, U_{\mu}\big]_N
  \Big\}
\end{equation}
%
where the probability of getting one probability given another follows the
detailed balance equation
%
\begin{equation}
  \frac{%
    P\big([\,]_{n\hphantom{-1}} \big| [\,]_{n-1}\big)
  }{%
    P\big([\,]_{n-1} \big| [\,]_{n\hphantom{-1}}\big)
  } = e^{\mathcal{S}[\,]_{n-1} - \mathcal{S}[\,]_n} \,,
\end{equation}
%
which will converge to the correct distribution as long as the chain is
\emph{ergodic}, i.e. will reach any field configuration in finite (albeit
arbitrarily long) time.

\subsection{Observables}

Using this process to select a finite set of configurations, one can
obtain an estimate for the expectation value of observables. 
Commonly named the \emph{ensemble average}, it is given by
%
\begin{equation}
  \bar{O}\big[\psi, \bar{\psi}, U_{\mu}\big] = \frac{1}{N} \sum_{n = 1}^N O\big( [\,]_n \big).
\end{equation}
%
If the set of configurations behaves as described in the previous section, it
converges to the true average, \meqref{eq:expectation_value}, when $N \to
\infty$.

We want an estimate of the error for any realisable number of configurations.
This information is encoded in the \emph{autocorrelation function}
%
\begin{equation}
  C_{O}(m) = \big\langle O\big([\,]_n\big) O\big([\,]_{n+m}\big)\big\rangle_n
    - \big\langle O\big([\,]_n\big)\big\rangle_n^2\,,
\end{equation}
%
which quantizes how correlated consecutive measurements are. Since the $n+1$'th
configuration depends on the $n$'th configuration through the Markov chain,
$C_O(m) \neq 0$ for $m > 0$. But it is how quickly this function decreases that
determine how good the estimate for the average is. An important quantity for
this is the \emph{integrated autocorrelation time}
%
\begin{equation}
  \tau_O = \frac{1}{2} \sum_{m=-\infty}^{+\infty} \frac{C_O(m)}{C_O(0)}.
\end{equation}
%
One can show that approximately $2 \tau_O$ of the consecutive measurements will
be correlated, meaning that it is advantageous to sample only the
observable at this interval instead of for every new configuration
\citep[sect. 7.1.3]{montvay1997quantum}.
